CodeBook for Getting and Cleaning Data Project 1 
========================================================
This document describes the variables, the data, and any transformations or work
performed to clean up the data.

Variables
-------------
Quoting from the original feature description: 

The features selected for this database come from the accelerometer and gyroscope 3-axial raw signals tAcc-XYZ and tGyro-XYZ. These time domain signals (prefix 't' to denote time) were captured at a constant rate of 50 Hz. Then they were filtered using a median filter and a 3rd order low pass Butterworth filter with a corner frequency of 20 Hz to remove noise. Similarly, the acceleration signal was then separated into body and gravity acceleration signals (tBodyAcc-XYZ and tGravityAcc-XYZ) using another low pass Butterworth filter with a corner frequency of 0.3 Hz. 

Subsequently, the body linear acceleration and angular velocity were derived in time to obtain Jerk signals (tBodyAccJerk-XYZ and tBodyGyroJerk-XYZ). Also the magnitude of these three-dimensional signals were calculated using the Euclidean norm (tBodyAccMag, tGravityAccMag, tBodyAccJerkMag, tBodyGyroMag, tBodyGyroJerkMag). 

Finally a Fast Fourier Transform (FFT) was applied to some of these signals producing fBodyAcc-XYZ, fBodyAccJerk-XYZ, fBodyGyro-XYZ, fBodyAccJerkMag, fBodyGyroMag, fBodyGyroJerkMag. (Note the 'f' to indicate frequency domain signals). 

These signals were used to estimate variables of the feature vector for each pattern: '-XYZ' is used to denote the 3-axial signals in the X, Y and Z directions.

Only the measurements on the mean and the standard deviation for each measurement are in scope of this analysis. 
```{r, feature_extraction, echo=TRUE}
# determine which features are means and standard deviations
feature <- read.table("./UCI HAR Dataset/features.txt")
indx <- feature[grepl("-(mean|std)\\(\\)",feature$V2,feature),]$V1
```
Intially, the entire set of features is loaded which contains `r nrow(feature)`
features and all variables having *std* or *mean* in its names are extracted 
from it, which narrows it down to `r length(indx)` features enumerated below.

Names are normalised by setting it to lower case and removing any brackets.
```{r, name_normalisation, echo=TRUE}
# normalise names: lowercase and removal of brackets
featurenames <- gsub("[()]","",tolower(feature[indx,]$V2))
```

The final list of `r length(indx)` originate from the following list of features:
* tbodyacc (6 variables: mean and std per x,y,z)
* tgravityacc (6 variables: mean and std per x,y,z)
* tbodyaccjerk (6 variables: mean and std per x,y,z)
* tbodygyro (6 variables: mean and std per x,y,z)
* tbodygyrojerk (6 variables: mean and std per x,y,z)
* tbodyaccmag (2 variables: mean and std)
* tgravityaccmag (2 variables: mean and std)
* tbodyaccjerkmag (2 variables: mean and std)
* tbodygyromag (2 variables: mean and std)
* tbodygyrojerkmag (2 variables: mean and std)
* fbodyacc (6 variables: mean and std per x,y,z)
* fbodyaccjerk (6 variables: mean and std per x,y,z)
* fbodygyro (6 variables: mean and std per x,y,z)
* fbodyaccmag (2 variables: mean and std)
* fbodybodyaccjerkmag (2 variables: mean and std)
* fbodybodygyromag (2 variables: mean and std)
* fbodybodygyrojerkmag (2 variables: mean and std)

Naming convention is ***{normalised original name}-(mean|std)-[xyz]*** for spatial
features and ***{normalised original name}-(mean|std)*** for non-spatial features.

Transformation
-----------------------

Once relevant features are determined, they are loaded from two data sets
and merged together using the following logic.
```{r}
readUCIHARDataset <- function(kind, names, cols) {
  dir <- paste0("./UCI HAR Dataset/",kind)
  dt <- data.table(read.table(paste0(dir,"/X_",kind,".txt"),colClasses=cols))
  setnames(dt, names)
  dt[, activity := read.table(paste0(dir,"/y_",kind,".txt"))]
  dt[, subject := read.table(paste0(dir,"/subject_",kind,".txt"))]
}

# import only relevant data from test and train datasets
cols <- rep("NULL",length(feature$V1))
cols[indx] <- "numeric"

# read only relevant variables from test and train datasets
testdt <- readUCIHARDataset("test",featurenames,cols)
traindt <- readUCIHARDataset("train",featurenames,cols)

# merge test with train
dt <- rbindlist(list(testdt,traindt))
```

Before the final computation we are replacing activity identifiers with 
labels read from a separate dictionary file.
```{r}
# replace activity number with label
activityname <- read.table("./UCI HAR Dataset/activity_labels.txt")
dt[, activity := as.factor(activityname[activity,]$V2) ]
```

Finally, a tidy data set is a projection of our data set through subject 
and activity aggregated by a mean function.
```{r}
# compute tidy dataset: reduce: mean(.SD) per subject per activity
dtt <- dt[,lapply(.SD,mean),by=list(subject,activity)]
write.table(dtt,"tidy.txt",row.names=FALSE)
```
